---
title: "Report"
format: html
editor: visual
---

```{r}
#| label: set-up
#| eval: true
#| echo: false

library(targets)

```

# Introduction

Climate change is expected to have large impacts on both marine [@perry2005] and terrestrial [@topper2018] ecosystems in Norway.

# Methods

## Field methods

The experiment was carried out in the high Arctic, close to Longyearbyen, Svalbard.
In 2018, sites were established along two elevational gradients with and without nutrient input from nesting sea birds [@vandvik2023plant].

## Data Analysis

Trait data are archived at https://osf.io/smbqh

Data were analysed in R version `r paste(R.version$major, R.version$minor, sep = ".")` [@R], using tidyverse for data manipulation and plots [@tidyverse].


## Results

```{r}
#| label: tbl-mean-area
#| tbl-cap: Mean leaf area per treatment.

tar_read(traits)
traits |> 
  group_by(Gradient) |> 
  summarise(n = n(), mean_area = mean(Value)) |> 
  gt() |> 
  cols_label(n = "Number of plants", mean_area = "Leaf area")
  
```



```{r}
#| label: fig-trait-figure
#| eval: false
#| echo: true


```


## References
