---
title: "Effect of climate change on Bistorta vivipara traits on Svalbard"
author: "Angelina Jonson"
format: html
editor: visual
date: today
bibliography: references.bib
csl: nordic-journal-of-botany.csl
execute:
  echo: false
---

```{r}
#| label: load-packages
#| message: false
library(tidyverse) # for ggplot2, dplyr, readr etc
library(here)      # for help finding files
library(lubridate) # for handling dates & times
library(gt)        # for tables
library(broom)     # for extracting results from models

# default plot theme
theme_set(theme_bw())

# default number of digit
options(digits = 2)
```

# Introduction

Climate change is expected to have large impacts on both marine [@perry2005] and terrestrial [@topper2018] ecosystems in Norway.

# Methods

## Field methods

The experiment was carried out in the high Arctic, in Endalen (78°11'N, 15°45'E), approximately four kilometres east of Longyearbyen, Svalbard, at 80-90 m elevation. The experiment was established in 2002 in three different habitats which are located in the south--southeast--facing hillside of the valley. The habitats differ in vegetation composition and the time of snowmelt (and hence the length of the growing season). The habitats include a relatively dry Dryas heath with thin snow cover (ca. 10 cm) and early snowmelt, a mesic Cassiope heath habitat with intermediate snow depth and melting date, and a moist snowbed community with deep snow (\> 100 cm) and late snowmelt.

In 2001, ten plots (75x75 cm) were selected in each of the three habitats. Half of the plots were randomly assigned to the warming treatment in 2002 (open-top chambers, OTC) and the other half to control (CTL).

## Data Analysis

Trait data are archived at https://osf.io/smbqh

Data were analysed in R version `r paste(R.version$major, R.version$minor, sep = ".")` [@R], using tidyverse for data manipulation and plots [@tidyverse].

# Results

```{r}
#| label: load-data
#| include: false
traits <- read_delim(
    file = here("data/PFTC4_Svalbard_2018_ITEX_Traits.csv")) 

# Filter out the required species
bistorta <- traits |>
  filter(Taxon == "bistorta vivipara")

```

```{r}
#| label: model

bistorta_height <- bistorta |> 
  filter(Trait == "Plant_Height_cm")

mod <- lm(Value ~ Treatment, data = bistorta_height)
mod_summary <- tidy(mod)

```

In total, we measured the height of `r nrow(bistorta_height)` Bistorta vivipera plants across the different habitats (@tbl-height-means). The Bistorta vivipera in the OTC plots is `r mod_summary$estimate[2]` cm ± `r mod_summary$std.error[2]` taller than in the control plots.

```{r}
#| label: tbl-height-means
#| message: false
#| fig-width: 5.2
#| tbl-cap: Mean height and number of individuals per treatment and habitat type (CH = Cassiope heath, DH = Dryas heath and SB = Snowbed).

bistorta_height |> 
  group_by(Treatment, Site) |> 
  summarise(n = n(), mean_height = mean(Value)) |> 
  gt() |> 
  cols_label(n = "Number of plants", Site = "Habitat", mean_height = "Height cm")
  
```

```{r}
#| label: fig-height-effect
#| fig-cap: Mean plant height in control (CTL) and warm (OTC) plots.
#| fig-width: 5.2

ggplot(bistorta_height, aes(x = Treatment, y = Value)) +
  geom_boxplot(fill = "grey80") +
  labs(x = "Treatment", y = "Height cm")
```

```{r}
#| label: fig-thickness-effect
#| fig-width: 5.2
#| fig-cap: Mean leaf thickness in control (CTL) and warm (OTC) plots.

leaf_thickness <- bistorta |> 
  filter(Trait == "Leaf_Thickness_mm")

ggplot(leaf_thickness, aes(x = Treatment, y = Value)) +
  geom_boxplot(fill = "grey80") +
  labs(x = "Treatment", y = "Leaf Thickness mm")
```

We measured the leaf thickness of `r nrow(leaf_thickness)` Bistorta vivipera plants across the different habitats.

```{r}
#| label: tbl-thickness-means
#| message: false
#| tbl-cap: Mean leaf thickness per treatment and habitat type (CH = Cassiope heath, DH = Dryas heath and SB = Snowbed).

leaf_thickness |> 
  group_by(Treatment, Site) |> 
  summarise(mean = mean(Value),
            sd = sd(Value),
            n = n()) |> 
  gt() |> 
  cols_label(Site = "Habitat", 
             mean = "Mean",
             sd = "Standard deviation",
             n = "Number") |> 
  fmt_number(columns = c(mean, sd, n)) |> 
  tab_spanner(label = "Leaf thickness", columns = c(mean, sd))

```
